### 2015-05-15 whoop.tex
fixes

\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
%
SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to reuse robust existing techniques and focus instead on verification efforts.

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentations and optimizations.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. However, it is relatively easy to enhance our tool with knowledge of other locking primitives. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. The are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.  The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.  We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.  If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as post-conditions for $\instrument{f}{i}$.  These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.  Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.  We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded by Houdini.  The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

@@

First, we enriched \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision. The Linux kernel can serialize calls to entry points, thus forcing them to run sequentially with each other. For example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock. We discovered this when \whoop reported many races between a number of networking entry points of the r8169 driver (see \S\ref{evaluation}); when we investigated the source of these races, we found out that these entry points could not race in reality because of RTNL. \whoop exploits this knowledge and does not create pairs for entry points that are mutually serialized by the kernel. This is an ongoing manual effort: the more drivers we study, the more such properties we discover to make \whoop more precise. Currently, this information is hard-coded inside \whoop. In the future, we plan to provide a configuration file where the user can denote such kernel-specific information.

@@

Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time. This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a global flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks. Because \whoop over-approximates read accesses to shared variables, and thus this flag, it can falsely report a pair as racy. However,~\cite[p.\ 173]{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).

>>>>

\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation and, hence, we do not need to keep track of them separately.
%
SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. Thus, \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts.

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentations and optimizations.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. However, it is relatively easy to enhance our tool with knowledge of other locking primitives. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. There are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.  The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.  We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.  If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as post-conditions for $\instrument{f}{i}$.  These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.  Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.  We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculate candidates that are later deemed to be incorrect, and thus discarded by Houdini.  The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

@@

First, we enriched \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision. The Linux kernel can serialize calls to entry points, thus forcing them to run in sequence instead of an interleaved manner. For example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock. We discovered this when \whoop reported many races between a number of networking entry points of the r8169 driver (see \S\ref{evaluation}); when we investigated the source of these races, we found out that these entry points could not race in reality because of RTNL. \whoop exploits this knowledge and does not create pairs for entry points that are mutually serialized by the kernel. This is an ongoing manual effort: the more drivers we study, the more such properties we discover to make \whoop more precise. Currently, this information is hard-coded inside \whoop. In the future, we plan to provide a configuration file where the user can denote such kernel-specific information.

@@

Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time. This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a global flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks. Because \whoop over-approximates read accesses to shared variables, and thus this flag, it can falsely report a pair as racy. However,~\cite[p.\ 173]{corbet2005linux} advises against serializing drivers in this way, as it ``inhibits user ingenuity'' (e.g.\ a user might expect that a driver can be accessed concurrently for performance).
